Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-03-29 src/tex/ms.tex
[showyourwork] overleaf sync

\subsection{Ocean Glint}
\cite{2001Natur.412..885F} noted the potential for time-resolved reflected light photometry to identify water oceans on the surface of an exoplanet.  \cite{2008Icar..195..927W} performed simulations of scattered light from directly imaged terrestrial exoplanets. 
%to demonstrate that phase-resolved photometry and polarization of a directly-imaged exoplanet would betray the presence of large surface oceans. 
They showed that oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases. %Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but
Specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., Titan's methane lakes;][]{2010GeoRL..37.7104S}. %\citet{lustig2019} explored mapping exoplanets at crescent phases using ocean glint.
Specular reflection appears at similar phase angles in both unpolarized and polarized light and its peak orbital phase can be extended by atmospheric winds, which produce waves and hence increase the size of the glint spot \citep{2011Icar..211..722B,kopparla2018}.
%\cite{2006astro.ph.10518M} quantified how photometric and polarimetric scattered light orbital phase variations might betray the presence of oceans on terrestrial exoplanets.  
\citet{Takahashi2021} used polarimetric Earthshine measurements to detect the presence of the ocean, proving that the method works in practice for an unresolved world.
%\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases.

@@

% Move to Clouds?
In practice, clouds complicate the detection of ocean glint because they obscure the underlying surface and can act as strong forward scatterers, mimicking ocean glint. \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to forward scattering from clouds, with a relatively small contribution from ocean glint.
\cite{Zugger_2010} simulated photometric and polarized phase variations for Earth-like planets including the effects of partial cloud coverage in suppressing the Rayleigh scattering and ocean glint polarization. Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they again find that oceans are detectable and distinguishable from other sources of polarization. To suppress the effects of Rayleigh scattering, glint observations can be made at longer wavelengths
\citep{Zugger_2011}.

@@

\subsection{Polarized Phase Curves}
\textcolor{blue}{Recap all of polarized phase curve work, including stuff not about ocean glint. Introduce the polarization signals of Rayleigh scattering, rainbow and glory.}
%compare to features in nonpol (strength rainbow); current capabilties?, errors from 3 papers,

@@

 Phase curves of planets are more complex than simple Lambertian curves.
 %
 The phase curve of an edge on system in unpolarized light is dominated by the reflectance and backscatter of the surface, clouds, and atmosphere.
 %
 Though this behaviour is not Lambertian, the planet will, in general, get brighter at fuller phases (small scattering angles) with a peak at full phase from cloud backscatter \citep[see, for example,][]{kopparla2018}.
 %
 In cases where surface liquid (sea) is present, the specular reflection can create a notable addition to the phase curve at broad scattering angles near crescent phase \citep{Robinson_2010}.
 %
 Exoplanet phase curves can also contain a rainbow signature at gibbous phases (moderately narrow scattering angles), though this is a small feature in comparison to the reflectance.

 In polarized light, additional structure in the phase curve can enable phenomena like ocean glint, rainbows, and the Rayleigh scattering to be more easily distinguished.  Notably, in the polarized phase curve the rainbow may be an especially strong feature and the maximum reflectance signal (from Rayleigh scattering) occurs

@@

In the non-eccentric case (i.e., with a circular orbit and where the coronagraphic mask centred on the star), minimum ($\varphi_\mathrm{min}$) and maximum ($\varphi_\mathrm{max}$) scattering phase angles accessible are symmetric about 90 degrees. 
%
We define the angle $\Delta \varphi$ such that: 
\begin{equation}
    \label{eq:Delta_phi}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2(\varphi_\mathrm{max} - 90) 
    =  2(\varphi_\mathrm{min} + 90)
\end{equation}

The on-sky projected planet distance, $r_\mathrm{proj}$ (in \si{\au}) is shown in \cref{fig:scattering-angle} assuming a non-eccentric circular orbit with a semi-major $a$ (in \si{\au}) at a distance $d_*$ (in \si{\parsec}) from the observer. 
%
The scattering phase angle $\varphi$ is this set up follows the equation,
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a}
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:

\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a}
\end{equation}

\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The on-sky projected planet distance, $r_\mathrm{proj}$, of a planet with a semi-major $a$ at a distance $d_*$. The on-sky angular separation as viewed by the observer is $\delta$. The scattering phase angle, $\varphi$, is the angle between the line of sight to the star and the line connecting the star and planet.  
    }
    \label{fig:scattering-angle}
\end{figure}

@@

Using \cref{eq:scattering_angle}, we deduce the equation for inclined circular orbits: 
\begin{equation}
\label{eq:Delta_phi_max}
    \Delta \varphi = 
    \begin{cases}
        2 \cdot i & \textrm{for} \cos(i) > \dfrac{\mathrm{IWA}\cdot d_* }{a}
  \\ 
        2 \cdot \cos^{-1}\left(\dfrac{\mathrm{IWA}\cdot d_* }{a}\right)  & \textrm{for} \cos(i) < \dfrac{\mathrm{IWA}\cdot d_* }{a}
    \end{cases}\,.
\end{equation}

>>>>

\subsection{Ocean Glint, Cloud Glory and Rainbows}
\cite{2001Natur.412..885F} noted the potential for time-resolved reflected light photometry to identify water oceans on the surface of an exoplanet.  \cite{2008Icar..195..927W} performed simulations of scattered light from directly imaged terrestrial exoplanets. 
%to demonstrate that phase-resolved photometry and polarization of a directly-imaged exoplanet would betray the presence of large surface oceans. 
They showed that oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases. %Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but
Specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., Titan's methane lakes;][]{2010GeoRL..37.7104S}. %\citet{lustig2019} explored mapping exoplanets at crescent phases using ocean glint.
Specular reflection appears at similar phase angles in both unpolarized and polarized light and its peak orbital phase can be extended by atmospheric winds, which produce waves and hence increase the size of the glint spot \citep{2011Icar..211..722B,kopparla2018}.
%\cite{2006astro.ph.10518M} quantified how photometric and polarimetric scattered light orbital phase variations might betray the presence of oceans on terrestrial exoplanets.  
\citet{Takahashi2021} used polarimetric Earthshine measurements to detect the presence of the ocean, proving that the method works in practice for an unresolved world.

%\noindent {\color{Green}<Modifications by Victor Trees>}

%\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases.

@@

% Move to Clouds?
In practice, clouds complicate the detection of ocean glint because they obscure the underlying surface and can act as strong forward scatterers, mimicking ocean glint in the total  reflected flux (polarized plus unpolarized). \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to forward scattering from clouds, with a relatively small contribution from ocean glint.

\cite{Zugger_2010} simulated photometric and polarized phase variations at visible wavelengths for Earth-like planets including the effects of %partial cloud coverage
Rayleigh scattering and a horizontally homogeneous cloud cover. They concluded that detection of an ocean glint signature in polarization is severely limited by the presence of clouds. %Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they %again 
%find that oceans are detectable and distinguishable from other sources of polarization. 
%{\color{Green} 
\citet{treesstam2019} simulated the total and polarized fluxes and the degree of polarization of light reflected by ocean planets with patchy clouds. They found that the alternation of cloud-free and cloud-covered glint, as the planet rotates and its cloud cover changes in time, can result in strong fluctuations in the degree of polarization, hence resulting in noisier phase curves. Nonetheless, polarization phase curves of ocean planets at optical and near infrared wavelengths intersect \citep{treesstam2019}, providing a possible avenue to distinguish between an exo-ocean and forward scattering clouds.%}%}

%{\color{Green}
The Rayleigh scattering optical thickness of a planet's atmosphere is smaller in the near-infrared (NIR) than in the visible, which improves the visibility of the glint \citep{Zugger_2011}. NIR phase measurements also minimize the latitude--albedo effect, which could otherwise be a false-positive for photometric brightening at crescent phases due to glint \citep{2012ApJ...752L...3C}.

Phase variations of polarized light could also reveal other non-Lambertian cloud scattering phenomena. The degree of polarization of light scattered by spherical cloud droplets strongly depends on the scattering angle \citep[see][]{hansentravis1974}, and consequently, on the phase angle for an Earth-like exoplanet with liquid water clouds \citep[see][]{stam2008}. The sign of the degree of polarization (i.e. the direction of polarization) changes sign, and show peaks at the primary and secondary rainbow angles  \citep{bailey2007,2018MNRAS.480.1613B}. A measurement of those rainbows would be a strong indicator of spherically shaped particles in the exoplanet's atmosphere. The precise phase angle locations in the the degree of polarization of the rainbow peaks depend on wavelength \citep[see Fig. 5 of][]{trees2022}, the particle size ...  and on the type of liquid \citep{bailey2007}. A famous example... Venus ... %A famous example of using polarimetry to characetrize clouds within our Solar System is the characterization of the main cloud deck of Venus.  \citep{hansenhovenier1974} derived that the main cloud deck of Venus consists of a 75\% sulfuric acid solution.
%
... glory ...\citep{2014A&A...566L...1G} ... and halos ... ice clouds ...

 \textbf{Victor/Kim: please complete this paragraph with a few sentences describing what one learns from rainbow and glory, ideally backed with good references!}

@@

%\subsection{Polarized Phase Curves}
%\textcolor{blue}{Recap all of polarized phase curve work, including stuff not about ocean glint. Introduce the polarization signals of Rayleigh scattering, rainbow and glory.}
%compare to features in nonpol (strength rainbow); current capabilties?, errors from 3 papers,

@@

 %Phase curves of planets are more complex than simple Lambertian curves.
 %
 %The phase curve of an edge on system in unpolarized light is dominated by the reflectance and backscatter of the surface, clouds, and atmosphere.
 %
 %Though this behaviour is not Lambertian, the planet will, in general, get brighter at fuller phases (small scattering angles) with a peak at full phase from cloud backscatter \citep[see, for example,][]{kopparla2018}.
 %
 %In cases where surface liquid (sea) is present, the specular reflection can create a notable addition to the phase curve at broad scattering angles near crescent phase \citep{Robinson_2010}.
 %
 %Exoplanet phase curves can also contain a rainbow signature at gibbous phases (moderately narrow scattering angles), though this is a small feature in comparison to the reflectance.

 %In polarized light, additional structure in the phase curve can enable phenomena like ocean glint, rainbows, and the Rayleigh scattering to be more easily distinguished.  Notably, in the polarized phase curve the rainbow may be an especially strong feature and the maximum reflectance signal (from Rayleigh scattering) occurs

@@

In the non-eccentric case (i.e., with a circular orbit and where the coronagraphic mask centred on the star), minimum ($\varphi_\mathrm{min}$) and maximum ($\varphi_\mathrm{max}$) scattering phase angles accessible are symmetric about 90 degrees. 
%
We define the angle $\Delta \varphi$ such that: 
\begin{equation}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2(\varphi_\mathrm{max} - 90^\circ) 
    =  2(90^\circ-\varphi_\mathrm{min} ) \notag
\end{equation}
% \begin{equation}
%     \label{eq:Delta_phi}
%     \Delta \varphi 
%     = \varphi_\mathrm{max} - \varphi_\mathrm{min}
%     =  2(\varphi_\mathrm{max} - 90) 
%     =  2(\varphi_\mathrm{min} + 90)
% \end{equation}

% {\color{Green} Victor Trees: Do you mean:
%INDEED, THERE MUST HAVE BEEN A TYPO, THX!
% %
% \begin{equation}
%     \Delta \varphi 
%     = \varphi_\mathrm{max} - \varphi_\mathrm{min}
%     =  2(\varphi_\mathrm{max} - 90^\circ) 
%     =  2(90^\circ-\varphi_\mathrm{min} ) \notag
% \end{equation}
%
% }

The on-sky projected planet distance, $r_\mathrm{proj}$ (in \si{\au}) is shown in \cref{fig:scattering-angle} assuming a non-eccentric circular orbit with a semi-major $a$ (in \si{\au}) at a distance $d_*$ (in \si{\parsec}) from the observer. 
%
The scattering phase angle $\varphi$ in this set up follows the equation,
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a}
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:
%
\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a}
\end{equation}
 %
\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The on-sky projected planet distance, $r_\mathrm{proj}$, of a planet with a semi-major $a$ at a distance $d_*$. The on-sky angular separation as viewed by the observer is $\delta$. The scattering phase angle, $\varphi$, is the angle between the line of sight to the star and the line connecting the star and planet.  
    }
    \label{fig:scattering-angle}
\end{figure}

@@

Using \cref{eq:scattering_angle}, we deduce the equation for inclined circular orbits: 
\begin{equation}
\label{eq:Delta_phi_max}
    \Delta \varphi = 
    \begin{cases}
        2 i & \textrm{for} \cos(i) > \frac{\mathrm{IWA}\; d_* }{a}
  \\ 
        2  \cos^{-1}\left(\dfrac{\mathrm{IWA}\cdot d_* }{a}\right)  & \textrm{for} \cos(i) < \frac{\mathrm{IWA}\; d_* }{a}
    \end{cases}\,.
\end{equation}
